Precise spiking motifs in neurobiological and neuromorphic data

### 2022-11-11 manuscript.tex
chip review

Artificial intelligence has made huge advances in the past decades, now deep learning algorithms outperform humans at complex tasks like natural image recognition or abstract strategy board games. Machine learning algorithms still suffer from adversarial attacks or a lack of generalization capacity but their main weakness, compared to biological neural networks, rely on their poor energy efficiency. Neuromorphic engineering intends to mimic the neural bases of communication with a wide variety of technics, from strictly analog circuits to software-based neuromorphic systems, and to develop tools improving the capacities of current artificial intelligence (\cite{roy2018toward}). Because the reduced energy consumption of biological networks can be explained by the use of spikes and asynchronous responses to exchange information (ref), neuromorphic chips use this parallel and event-based representation to make energy efficient computations. This is also the case for Dynamic Vision Sensors (DVS) which provide a stream of binary asynchronous events signaling detectable changes in luminance. These devices, also named ``silicon retinas'' (see Figure \ref{fig:silicon_retina}), show great improvements in terms of memory allocation for the recording of a visual scene (ref). Another important distinction with classical Von Neumann architectures is the localized memory of this new type of chips. It is materialized by capacity of the physical connections between the processing units to store information. An example of such connection, directly inspired by synaptic plasticity, is the memristor~\citep{markovic_physics_2020}. Using these event-based computations as a building brick, neuromorphic engineering proposes new hardware designs perfect to simulate SNNs and use the full power of asynchronous computations observed in biological systems. Many connections can be drawn between neuromorphic engineering and computational neuroscience to aim at solving both research and technology challenges (zenke2021visualizing). Just as communication between neurons can be illustrated by a raster plot, information exchanges in neuromorphic hardware are described as event streams by an Address-Event Representation (AER). We will use indifferently \textit{events} or \textit{spikes} to describe the same binary and discrete processes and, in this section, we will extend the review of the use of precise spike timing to this emerging field.

%On one hand, neuromorphic engineering is inspired by biology to develop efficient algorithms and technology. For instance, DVS cameras are inspired by the retina, memristors are used to implement neuroplasticity and neuromorphic chips are designed to emulate neural elements in order to perform sparse and asynchronous computations as observed in the biological brain. Therefore, the neuromorphic field is interested in understanding neural circuitry to be able to reproduce such mechanisms. On the other hand, computational neuroscience uses mathematical models to describe and explain a wide range of cognitive processes. It focuses on biological plausibility to validate its models and in comparison with experimental data. As a result, both fields have a common interest in neural information processing and how these may implement the computations that happen in the brain (for a review, see~\cite{zenke2021visualizing}).

\subsection{On the importance of spatio-temporal information in event-based recordings}

%A time surface represents therefore a combination of an object and its motion, but both elements are not explicitly dissociated. This entangled information may need to be separated in order to reach optimum object categorization.

With the AER specification and their submillisecond temporal precision, DVS bring a new approach to the storage and use of visual information. From their generative model, these sparse events are markers of the dynamics of the visual scene captured by the sensor. The dynamics of the event streams have to be used to make sense of the recorded information and new algorithms are developed to solve efficiently classical computer vision tasks. In~\citep{lagorce_hots_2017}, \textit{time surfaces} are introduced as an event-driven 2D image of the delay between the last event recorded at the address of a pixel and the current time. An exponential decay is applied on this delay to obtain the analog values of the time surfaces, it gives more precision to represent recent events and offers an analogy with the LIF spiking neuron. Its a way to represent the local dynamics embedded in the event-based recordings, and with unsupervised learning on the event stream based on the cosine similarity they can capture repeating motifs on the input signals. Learned time surfaces can be used for object recognition~\citep{lagorce_hots_2017, sironi_hats_2018, maro_event_2020, grimaldi_homeostatic_2021,grimaldi_robust_2022}. Three dimensional convolutions in both space and time are another representation of the spiking motifs embedded in the event stream~\citep{ghosh_spatiotemporal_2019, grimaldi_learning_2022}. With their additional temporal dimension, they can capture multiple events on the same pixel address as long as they belong to the local temporal window. This representation is only limited by the time step used for the discretization of the signal, this factor defines the temporal precision of the representation. Other methods make direct use of the precise timing of events captured by the DVS to solve optical flow and time-to-contact challenges~\cite{benosman_event-based_2014, clady_asynchronous_2014, tschechne_bio-inspired_2014}, inferring depth~\citep{hidalgo_learning_2020}, feature detection and tracking~\citep{dardelet_event-by-event_2021}, motion segmentation~\citep{stoffregen_event-based_2019} or the simultaneous localization and mapping problem~\citep{kim_real-time_2016}. This non-exhaustive list of complex task solving are not directly linked to biological processes but show the potential of using the precise temporal resolution of retinomorphic cameras. By essence, these sensors offer a novel view of visual information processing and the use of spatio-temporal motifs of embedded in the event streams are essential to solve high level visual tasks.

@@

Remarkably, novel neuromorphic chips use a representation similar to that of real neurons~\citep{rasetto_challenges_2022}

% un exemple de polychronization
~\citet{wang_neuromorphic_2015} presented a hardware implementation of polychronous networks in which propagation delays are learned in a supervised manner, based on the expected firing time of the post-synaptic neuron. This paper proposes a supervised delay learning algorithm for spiking neurons with temporal encoding, in which both the weight and delay of a synaptic connection can be adjusted to enhance the learning performance.

@@

\citep{renner_sparse_2022}

\citep{sandamirskaya_neuromorphic_2022}

\citep{hussain_deltron_2012}
\citep{sandamirskaya_neuromorphic_2022}
\citep{sandamirskaya_neuromorphic_2022}

%https://ieeexplore.ieee.org/abstract/document/8123631

%https://ieeexplore.ieee.org/abstract/document/9622153

@@

%Sparse Vector Binding on Spiking Neuromorphic Hardware Using Synaptic Delays

@@

% solutions usuellement utilisées - surrogate gradient

% autres méthodes
Spatio-temporal kernels:~\citep{yu_stsc-snn_2022}

>>>>

Artificial intelligence has made huge advances in the past decades, now deep learning algorithms outperform humans at complex tasks like natural image recognition or abstract strategy board games. Machine learning algorithms still suffer from adversarial attacks or a lack of generalization capacity but their main weakness, compared to biological neural networks, rely on their poor energy efficiency. Neuromorphic engineering intends to mimic the neural bases of communication with a wide variety of technics, from strictly analog circuits to software-based neuromorphic systems, and to develop tools improving the capacities of current artificial intelligence (\cite{roy_towards_2019}). Because the reduced energy consumption of biological networks can be explained by the use of spikes and asynchronous responses to exchange information (ref), neuromorphic chips use this parallel and event-based representation to make energy efficient computations. This is also the case for Dynamic Vision Sensors (DVS) which provide a stream of binary asynchronous events signaling detectable changes in luminance. These devices, also named ``silicon retinas'' (see Figure \ref{fig:silicon_retina}), show great improvements in terms of memory allocation for the recording of a visual scene (ref). Another important distinction with classical von Neumann architectures is the localized memory of this new type of chips. It can be materialized by capacity of the physical connections between the processing units to store information~\citep{markovic_physics_2020}. An example of such connection, directly inspired by synaptic plasticity, is the memristor~\citep{rasetto_challenges_2022} for which the resistance value can be dynamically adjusted. Using these event-based computations as a building brick, neuromorphic engineering proposes new hardware designs perfect to simulate SNNs and use the full power of asynchronous computations observed in biological systems. This field of research is inspired by neuroscientific advances and computational formalisms to design innovative architectures and by reproducing such mechanisms it is interesting to study neural circuitry. Many connections can be drawn between neuromorphic engineering and computational neuroscience to aim at solving both research and technology challenges~\cite{zenke_visualizing_2021}. Just as communication between neurons can be illustrated by a raster plot, information exchanges in neuromorphic hardware are described as event streams by an Address-Event Representation (AER). We will use indifferently \textit{events} or \textit{spikes} to describe the same binary and discrete processes and, in this section, we will extend the review of the use of precise spike timing to this emerging field. We report different types of event-based sensing devices for sound~\cite{chan_aer_2007} and touch~\cite{haessig_event-based_2020} but first focus on neuromorphic cameras to review different methods using efficiently the dynamics of event streams. Then, we dedicate one subsection to the implementation of delay-based algorithms on neuromorphic chips.  
\subsection{On the importance of spatio-temporal information in silicon retinas}
With the AER specification and their submillisecond temporal precision, DVS bring a new approach to the storage and use of visual information. From their generative model, these sparse events are markers of the dynamics of the visual scene captured by the sensor. The dynamics of the event streams have to be used to make sense of the recorded information and new algorithms are needed to solve efficiently classical computer vision tasks. In~\cite{lagorce_hots_2017}, \textit{time surfaces} are introduced as an event-driven 2D image of the delay between the last event recorded at the address of a pixel and the current time. An exponential decay is applied on this delay to obtain the analog values of the time surfaces, it gives more precision to represent recent events and offers an analogy with the LIF spiking neuron. Its a way to represent the local dynamics embedded in the event-based recordings, and with unsupervised learning on the event stream based on the cosine similarity they can capture repeating motifs within the input signals. Learned time surfaces can be used for object recognition~\citep{lagorce_hots_2017, sironi_hats_2018, maro_event-based_2020, grimaldi_homeostatic_2021,grimaldi_robust_2022}. Three dimensional convolutions in both space and time are another representation of the spiking motifs embedded in the event stream~\citep{ghosh_spatiotemporal_2019, grimaldi_learning_2022, yu_stsc-snn_2022}. With their additional temporal dimension, they can capture multiple events on the same pixel address as long as they belong to the local temporal window. This representation is only limited by the time step used for the discretization of the signal, this factor defines the temporal precision of the representation.  Other methods make direct use of the precise timing of events captured by the DVS to solve optical flow and time-to-contact challenges~\citep{benosman_event-based_2014, clady_asynchronous_2014, tschechne_bio-inspired_2014}, inferring depth~\citep{hidalgo_learning_2020}, feature detection and tracking~\citep{dardelet_event-by-event_2021}, motion segmentation~\citep{stoffregen_event-based_2019} or the simultaneous localization and mapping problem~\citep{kim_real-time_2016}. This non-exhaustive list of complex task solving is not directly linked to biological processes but show the potential of using the precise temporal resolution of retinomorphic cameras. By essence, these sensors offer a novel view of visual information processing due to the asynchronous responses of the different pixels. With this type of signal, the use of spatio-temporal motifs embedded in the event streams are essential to solve high level visual tasks. 

@@

SNNs are considered the third generation of ANNs and, also for the asynchronous nature of computations, neuromorphic engineering is promising for the future of artificial intelligence. Both fields of research are characterized by local asynchronous computing units and the absence of a central clock and this is why the use of the temporal dimension in these types of architectures seems crucial. The previous section demonstrates the importance of using spatio-temporal motifs in event-based recordings, here we focus on the implementation of such event-based algorithms in neuromorphic hardware. Even if some useful SNNs simulators run on GPUs (brian, etc.. ref), such event-based computing technics show their advantages in terms of frugality and rapidity only on neuromorphic chips.

To our knowledge, the first neuromorphic circuit is the pulsed current-source synapse proposed by Carver Mead in 1989~\cite{mead_pulsed_1989}. It was implemented with transistors operating in the subthreshold domain and responded to asynchronous events but was not capable of discriminating two different spiking sequences with the same firing rate. Indeed, the postsynaptic membrane potential was increased by a step proportional to the input current but did not decreased in time like it can be observed in biological neurons. Then, electronics circuits became more and more bio-realistics and~\cite{bartolozzi_synaptic_2007} released the Diff-Pair Integrator (DPI) synapse that could reproduce the dynamics of the biological neurons. The DPI circuit could multiplex in time spikes from different sources and became a potential  ``silicon coincidence detector". Today many devices are good candidates for implementing event-based algorithms and use the AER, they can be divided intro 3 major categories: digital, analog and mixed analog/digital platforms. For a more complete review the reader can refer to~\cite{schuman_survey_2017}, here we site the most popular ones. SpiNNaker, Loihi and TrueNorth chips (ref) are widely used examples of digital hardware implementations. Compared to TrueNorth which exclusively implements a LIF neuron, SpiNNaker and Loihi offer some flexibility in terms of neuron model and allow for on-chip learning. This flexibility in the implementation comes at the cost of an increased energy consumption. Mixed-analog-digital systems were developed at the Stanford University: Neurogrid and Braindrop (ref). They are mostly used by computational neuroscientist to model brain activity with different levels of abstraction. BrainScaleS (FPGA, etc... voir la revue faite pour agileneurobot).

\citep{hussain_deltron_2012} reported an implementation of a delay-learning algorithm on an analog chip. Online learning on neuromorphic chips is still a challenge today and for this work, only the detection of spiking motifs was done on the analog architecture, training was performed digitally and based on the tempotron learning rule~\cite{gutig_tempotron_2006}. It

\note{parler un peu des chips qui existent? This poten-
tial advantage of spiking has motivated a number of neural
computing architectures such as IBM’s TrueNorth [16] and
University of Manchester’s SpiNNaker processor [17]. }
Oui commencer par décrire le développement de ces chips avec un historique, dire qu'elles sont

they made a Spike-Timing Neuromorphic Architecture with a dendritic arbor to implement the delays and a transfer function for the weights which is
well-suited for areas such as pattern recognition and natural
language processing. In this paper, we formally describe the
STPU, implement the STPU on a field programmable gate array,
and show measured performance data.\cite{hill_spike-timing_2017}

% un exemple de polychronization
~\citet{wang_neuromorphic_2015} presented a FPGA hardware implementation of polychronous networks in which propagation delays are learned in a supervised manner, based on the expected firing time of the post-synaptic neuron. This paper proposes a supervised delay learning algorithm for spiking neurons with temporal encoding, in which both the weight and delay of a synaptic connection can be adjusted to enhance the learning performance.

@@

However, the binding of spatially sparse representations is computationally expensive because it is not local to corresponding pairs of neurons as in VSAs with dense vectors. Here, we present the first implementation of a sparse VSA on spiking neuromorphic hardware, specifically Intel’s neuromorphic research chip Loihi. To reduce the cost of binding, a delay line and coincidence detection are used, trading off space with time. We show as proof of principle that our network on Loihi can perform the binding operation of a classical analogical reasoning task and discuss the cost of different sparse binding operations. The proposed binding mechanism can be used as a building block for VSA-based architectures on neuromorphic hardware. \citep{renner_sparse_2022}

review on neuromorphic computations for robotic\citep{sandamirskaya_neuromorphic_2022} (not accessible for the moment)

deltron is an implementation of delay learning based on membrane potential optimization/supervised learning to detect spatio-temporal patterns in spike train. Only delays no weights and supervision with timing of the patterns (not sure because they take Vmax and then compare with Vthres)\citep{hussain_deltron_2012}

aer ear a silicon cochlea with AER \citep{chan_aer_2007}

new analog VLSI synapse circuit (the DPI of section 2.6)
useful for implementing postsynaptic currents in neuromorphic VLSI net-
works of spiking neurons with biologically realistic temporal dynamics. \citep{bartolozzi_synaptic_2007} Several subthreshold synaptic circuit designs have been proposed (Mead, 1989; Lazzaro, 1994; Boahen, 1998; Fusi et al., 2000; Chicca, Indiveri,  Douglas, 2003; Shi  Horiuchi, 2004a; Gordon, Farquhar,  Hasler, 2004; Hynna  Boahen, 2006) covering a range of trade-offs between functionality and complexity of temporal dynamics versus circuit and layout size.
these properties in our VLSI devices because there is evidence that they play an important role in detecting coincidence between the presynaptic activity and postsynaptic depolarization for inducing long-term-potentiation (LTP) (Morris, Davis,  Butcher, 1990)

We apply delay-based reservoir computing as the information processing core, along with a novel training and labelling method. Different from the conventional ECG classification techniques, this computation model is a end-to-end dynamic system that mimics the real-time signal flow in neuromorphic hardware\cite{liang_neuromorphic_2021}

they implement on a neuromorphic chip the detection of spikes delay up to a 50 ns precision, simulation of the auditory cortex of the barn owl\cite{pfeil_neuromorphic_2013}

@@

@@

% solutions usuellement utilisées - surrogate gradient
for other section: A hypothesis for temporal coding of young and mature granule cells(https://www.frontiersin.org/articles/10.3389/fnins.2013.00075/full)
